### 2010-12-07 report/abstract.tex
much editing with sarah, diana

We herein propose a system...
\edit{Write abstract}

>>>>

% A brief summary of the report. This is not an introduction -
% it should be complete. Include 1-2 sentences for each of the items
% below, up to and including conclusions.

This paper presents an array of approaches to optimizing a quadrupedal
gait for forward speed.  We implement, test, and compare different
learning strategies including uniform and Gaussian random hill
climbing, policy gradient reinforcement learning\cite{kohl},
Nelder-Mead simplex\cite{nm}, new predictive methods based on linear
and support vector regression, and an evolved neural network
(HyperNEAT)\cite{2}.  We compare results to a baseline random search
method.  Many of the methods resulted in walks significantly faster
than previously hand-tuned gaits.

%Because the fastest learned walk
%was not significantly faster than the fastest randomly generated walk,
%we conjecture that the motion representation for the robot is more
%integral to forward speed than the learning algorithm.

\edit{fix citations}

Keywords: Learning Control, Walking Robots, Gait Learning, Quadruped

### 2010-12-07 report/report.tex
much editing with sarah, diana

We aimed to strike a middle ground between these approaches.  Our
motion generator did not relying on hand-tweaked gaits, but it did use
a parameterized gait which, by its nature, imposes some assumptions on
the answers produced.  From this model, we used machine learning to
design a gait for a quadruped robot.

This paper presents a comparison of varied learning methods to design
a gait for a quadruped robot. Using these methods we have created
walks that are several times faster than the original hand-tuned gait.
We invite readers with short attention spans to view a video of some
of our results online here:

@@

\section{Problem definition}
% Precisely define the problem you are addressing (i.e. formally specify
% the inputs and outputs). Elaborate on why this is an interesting and
% important problem.
We are testing several different learning methods to design a parametrized 
gait for a quadruped robot from the Cornell Computational Synthesis Lab. 
The metric for evaluation of the designed gait is speed. Given parameters 
for the current walk and the current time, our code outputs the position
of each of the robot's motors. A comparison and evaluation 
of the many different methods available for optimizing the gait of legged 
robots will be useful for future work on this challenging multidimensional 
control problem.

\section{Method}
% Describe in reasonable detail the algorithm you are using to address this 
% problem. A pseudo-code description of the algorithm you are using is 
% frequently useful. If it makes sense for your project, trace through a 
% concrete example, showing how your algorithm processes this example. The 
% example should be complex enough to illustrate all of the important aspects 
% of the problem but simple enough to be easily understood. If possible, 
% an intuitively meaningful example is better than one with meaningless 
% symbols.

We have a parameterized motion model that commands motors to positions
based on a sine wave, creating a periodic pattern. Parameters are: 
amplitude, wavelength, scale inner vs outer motors, scale left vs right 
motors, scale back vs front motors. Each strategy below attempts to choose
the best possible parameters for this motion model.

We implemented and tested 8 different learning strategies:
\begin{itemize}
\item \emph{Uniform random hill climbing}: Creates a random neighbor by
  randomly choosing one parameter to adjust, and changing it completely 
  randomly in UniformStrategy(). Then evaluates this neighbor by running 
  the robot with the newly chosen parameters. If this neighbor results in a
  longer distance walked then the previous best neighbor, we save this
  neighbor as the new best neighbor. We then repeat the whole process
  using the best neighbor as a base.
\item \emph{Gaussian random hill climbing}: Creates a random neighbor by
  changing each parameter randomly based on a normal distribution. Then
  evaluates this neighbor similarly to uniform random hill climbing.
\item \emph{N-dimensional policy gradient descent}: Estimates the policy
  gradient by evaluating \emph{t} randomly generated policies
  near an initial parameter vector. Then computes the average score for
  each parameter and adjusts the base policy according to the estimated
  gradient.
\item \emph{Random}: Randomly generates parameter vector in the allowable
  range. This strategy is used only as base line.
\item \emph{Simplex (Nelder-Mead) Method}: Creates a simplex with 6 vertices and tests the fitness of each vertex. The worst vertex is reflected over the centroid. If the reflected point is better than the second worst point and worse than the best point, then the reflected point replaces the worst. If the reflected point is better than the best point, the simplex is expanded in the direction of the reflected point. The better of the reflected and the expanded point replaces the worst point. If the reflected point is worse than the second worst point, then the simplex is contracted away from the reflected point. If the contracted point is better than the reflected point, the contracted point replaces the worst point. If the contracted point is worse than the reflected point, the entire simplex is shrunk \cite{NMWebsite}.
\item \emph{Linear regression}: To initialize, chooses five random points
and trains on them using least-squares. Then in a loop, takes a fixed-size
step in the direction of the gradient, and retrains on all points so far.
\item \emph{SVM regression}
\item \emph{Evolutionary Neural Network (HyperNEAT)\cite{2}}
\end{itemize}

@@

Various maching learning techniques have proven to be useful in finding
control policies for a wide variety of robots. Kohl and Stone\cite{1}
presented a policy gradient reinforcement learning approach for generating
a fast walk on legged robots. We experimented with this method to create
a walk for our robot (Policy Gradient Descent). Chernova and Velosa\cite{3}
took an evolutionary approach to this problem which we did not implement.
Zykov, Bongard, and Lipson\cite{4} describe the evolution of dynamic gaits 
on a physical robot requiring no prior
assumptions about the locomotion pattern beyond the fact that it should be
rhythmic.

@@

The quadruped robot has an on-board computer running Linux. The lower
level drivers are in C and the system is implemented in Python. Feedback about distance travelled is provided via an infrared LED and a Wii remote.

@@

\item \code{RunManager} class: Deals with all the details of running the robot,
  including choosing an initial parameter, running the robot multiple times, 
  tracking distance walked, and writing to the log file. Also includes the
  \code{explore\_dimensions} method.

\item \code{Strategy} class: The user has a choice between seven
  different learning strategies: uniform random hill climbing, Gaussian random
  hill climbing, N-dimensional policy gradient descent, random, 
  simplex, linear regression/prediction, and SVM regression/prediction.

\item \code{MotionModel} abstract base class: Several functions exist,
  but none are within a class framework. ... ??? ...

  \begin{itemize} 
  \item \code{SineModel}
  classes: Commands motors to positions based on a sine wave, creating
  a periodic pattern. Parameters are: amplitude, wavelength, scale
  inner vs outer motors, scale left vs right motors, scale back vs
  front motors. Currently a function, not a class. ... ??? ... There
  are also other motion models...

  \item Other derived motion model classes:
  Several versions of a SineModel exist, each with different
  parameters, though solely as functions.
  \end{itemize}

\item \code{WiiTrackClient} and \code{WiiTrackServer} classes and
  hardware: A Wii remote tracks the location of the robot through an
  infrared LED mounted on top of the robot. We used the CWiid library
  which is able to interface with the remote via bluetooth to
  determine the location of the infrared LED. This information is
  accessed through some functions we wrote and passed to the
  program. The program gets the robot's position at the beginning of
  each run and then again at the end. The program then calculates the
  net change in position and uses that as the distance
  walked. \edit{Jason: add details}.

@@

We controlled our experiments from a computer that was connected via a 
wireless ethernet to the robot. The robot collected data about distance
walked automatically on its own. If it walked outside of the Wii remote's
viewable area, it informed the user, so the only human intervention
required during an experiment involved moving the robot back inside the 
viewable area and restarting it and dealing with various mechanical failures,
which did not interrupt the learning process or result in the loss of data.

We evaluated the efficacy of a set of parameters by sending these
parameters to the robot and instructing it to walk for a certain length
of time. The robot always begins from the same position and returns to the
starting position at the end of the run in order to allow for uniformity
in measuring. More efficient parameters resulted in a faster gait, which
translated into a longer distance walked and a better score. After completing
an evaluation, the robot sent the resulting distance walked back to the
host computer and prepared itself for a new set of parameters to evaluate.

@@

The resulting gaits from our algorithms quickly outperformed the original
hand-coded walk designed for this robot. The fasted walk, for example, was
4 times faster.

@@

We have done runs of 
3 different initial parameter vectors for random hill climbing, 
Gaussian random hill climbing, policy gradient descent, random,
simplex, and linear regression. We developed several
gaits that were about 4 times faster than the original hand-coded gait.

\editbox{Pull out vectors A, B, and C into separate bulletted list, and
  make table, like below}

\begin{tabular}{|r|c|c|c|}
\hline
                                         & A       & B      & C      \\
\hline
Uniform Random Hill Climbing             & 11.37   & 9.44   & 2.69   \\
\hline
Gaussian Random Hill Climbing            & 3.10    & 13.59  & 13.40  \\
\hline
Policy Gradient Descent                  & 0.68    & 14.69  & 3.60   \\
\hline
Random                                   & 6.04    & 17.26  & 4.90   \\
\hline
Simplex (Nelder-Mead)                    & 8.51    & 13.62  & 14.83  \\
\hline
Linear Regression                        & 27.58   & 12.51  & 1.95   \\
\hline
SVM Regression                           & ???     & ???    & ???    \\
\hline
Evolutionary Neural Network (HyperNEAT)  & ???     & ???    & ???    \\
\hline
\end{tabular}

@@

\item For vector A, linear regression worked significantly better than the other algorithms, resulting in a gait (27.58 body lengths/minute) that walked over twice as fast as the next best gait (uniform random hill climbing - 11.37 body lengths/minute).

\item Vector B resulted in similarly performing gaits with all the algorithms. Random produced the best gait (17.26 body lengths/minute) and uniform random hill climbing produced the worst (9.44 body lengths/minute). The remaining algorithms each produced a gait around 13 body lengths/minute.

\item For vector C, simplex and Gaussian random hill climbing each produced a gait that substantially outperformed the other algorithms. Simplex resulted in a gait of nearly 15 body lengths/minute  and Gaussian random hill climbing produced a gait of just over 13 body lengths/minute, whereas the other algorithms returned gaits of less than 5 body lengths/minute.

@@

Based on the three trials, linear regression worked the best, followed by simplex. Gaussian random hill climbing performed about as well as random, but it is unclear whether Gaussian would continue to improve and become much better than random if allowed to run for more iterations. Uniform random hill climbing and policy gradient descent performed similarly.

@@

\edit{fix captions in this section too!}

\figvarp{vectorA}{.65}{Vector A.... more here....}{}

\figvarp{vectorB}{.65}{Vector B.... more here....}{}

\figvarp{vectorC}{.65}{Vector C.... more here....}{}

\figvarp{std_error}{1}{Standard error....}{}

@@

\edit{important points... talk about these more :)}

\begin{itemize}
\item  Since only three trials were tested with each algorithm and no algorithm consistently outperformed the others, there is a large standard error for each algorithm. For this reason, it is unclear if any algorithm is superior to another in this application. More trials with each algorithm would be necessary to reach a definitive ranking of the performance of the algorithms.

\item Each algorithm discovered at least one gait of over 10 body lengths/minute, including random. For this reason, we speculate that the motion model may be more critical than the learning algorithm.
\end{itemize}

@@

There are several additions we would make to flesh out our project. First,
the error margins for our runs was quite large, so reducing them would lead
to more conclusive data. It would also be insightful to run our algorithms
on different motion models to get a better idea of how well different
algorithms perform overall, and not just for our current motion model. We
also intend to experiment more with evolutionary algorithms/HyperNEAT. 
Some parameter vectors resulted in the robot turning, as opposed to it
walking long distances. This could be an interesting learning goal in
future projects. To these ends, we propose the following additions
and enhancements:

\begin{itemize}
\item More runs and/or longer runs
\item Different motion representations
\item Evolutionary algorithms/HyperNEAT\cite{2}
\item Learning how to turn
\end{itemize}

@@

\edit{make this more awesome}

Because the fastest learned walk was not
significantly faster than the fastest randomly generated walk, we
conjecture that the motion representation for the robot is more
integral to forward speed than the learning algorithm.

\section{Acknowledgments}
% List any people not on the team who helped you with your project: Your
% TA, other people you consulted with or had useful discussions (say in 
% a word or two what they did), people who proofread your report, and 
% any external code you used (libraries etc).
\begin{itemize}
\item Hod Lipson, Cornell Computational Synthesis Lab: adviser
\item Jim T\o rreson, University of Oslo: adviser
\item Juan Zagal, University of Chile: designed and printed robot and provided assistance
\item Jeff Clune, CCSL Red Couch: collaborated on HyperNEAT implementation/testing
\item Cooper Bills, Cornell University: assisted with Wii tracker development
\item Anshumali Srivastava, Cornell University: Teaching Assistant
\end{itemize}

\section{The Team}
% A table listing the names of the people who worked on the project, and who
% did what.
\begin{itemize}
\item Diana Hidalgo: 
\begin{itemize}
\item Simplex (Nelder-Mead)
\item \code{WiiTrackClient} and \code{WiiTrackServer} classes and hardware
\item Plotting and evaluating results
\end{itemize}
\item Sarah Nguyen
\begin{itemize}
\item Random
\item Uniform random hill climbing
\item Gaussian random hill climbing
\item N-dimensional policy gradient descent
\item Linear regression and prediction
\item \code{RunManager} and \code{Strategy} classes
\item Video editing
\end{itemize}
\item Jason Yosinski
\begin{itemize}
\item SVM regression and prediction
\item Evolutionary Neural Network (HyperNEAT)
\item \code{WiiTrackClient} and \code{WiiTrackServer} classes and hardware
\item \code{Robot}, \code{RunManager}, and \code{Strategy} classes
\item \code{Parameterized Motion Model} abstract base class
\item \code{SineModel} classes
\item \code{explore\_dimensions}
\item Plotting and evaluating results
\end{itemize}
\end{itemize}

@@

\begin{itemize}
\item \code{optimize.py}: Determines a strategy to try and runs the robot
  with that strategy.
\item \code{RunManager.py}: Deals with all the details of running the robot,
  including choosing an initial parameter, running the robot multiple times, 
  tracking distance walked, and writing to the log file. Also includes the
  \code{explore\_dimensions} method.
\item \code{Strategy.py}: Contains all the different possible strategies, which
  will be passed as objects in \code{optimize.py}. 
\item \code{Robot.py}: Implements the \code{Robot} class, described in
  \secref{implement}.
\item \code{SineModel.py}: Implements a sine based motion model,
  described in \secref{implement}.
\item \code{Motion.py}: Motion helper functions.
\item \code{WiiTrackServer.py}: Broadcasts the position of the infrared LED.
\item \code{WiiTrackClient.py}: Connects to the WiiTrackServer to get the current position information.
\end{itemize}

>>>>

We aimed to strike a middle ground between these approaches.  Our
motion generator did not rely on hand-tweaked gaits, but it did use
parameterized gaits which, by their nature, impose some assumptions on
the answers produced.  We then used machine learning to design gaits
for a quadruped robot with these models.  This paper presents a
comparison of the different learning methods implemented.  Most
methods created walks that are several times faster than the original
hand-tuned gait.  We invite readers with short attention spans to view
a video of some of our results online here:

@@

\section{Problem definition}
% Precisely define the problem you are addressing (i.e. formally specify
% the inputs and outputs). Elaborate on why this is an interesting and
% important problem.

We are testing several different learning methods to design a
parametrized gait for a quadruped robot from the Cornell Computational
Synthesis Lab.

The output each of the learning algorithms is a function of time,
$f(t)$, that outputs a vector of commanded motor positions.  This
function is generated using a parametrized motion model, described in
\secref{motion}.

The robot executes these commands and measures its change in location
using the tracking system described in \secref{implement}.  The input
to the learning algorithms is this measured displacement, which the
algorithms attempt to maximize. This displacement is measured for each
gait over a constant length run, usually 12 seconds.

A comparison and evaluation of the many different methods available
for optimizing the gait of legged robots will be useful for future
work on this challenging multidimensional control problem.

\section{Method}
\seclabel{method}

% Describe in reasonable detail the algorithm you are using to address
% this problem. A pseudo-code description of the algorithm you are
% using is frequently useful. If it makes sense for your project,
% trace through a concrete example, showing how your algorithm
% processes this example. The example should be complex enough to
% illustrate all of the important aspects of the problem but simple
% enough to be easily understood. If possible, an intuitively
% meaningful example is better than one with meaningless symbols.

We use several parameterized motion models that command motors to
positions based on a sine wave, creating a periodic pattern.  While we
investigated several models, for the bulk of our experiments, we used
a model whose five parameters are: amplitude, wavelength, scale inner
vs outer motors, scale left vs right motors, scale back vs front
motors. Each strategy below attempts to choose the best possible
parameters for this motion model.

We implemented and tested 8 different learning strategies.  All
strategies except for the HyperNEAT method\cite{clune} were
constrained to pick parameters from within predetermined ranges.

\item \emph{Random}: This method randomly generates parameter vectors
  in the allowable range. This strategy is used only as baseline.

\item \emph{Uniform random hill climbing}: This method begins by
  selecting a single random parameter vector.  Subsequent iterations
  generate a neighbor by randomly choosing one parameter to adjust and
  replacing it with a new value chosen with uniform probability in the
  allowable range for that parameter. The neighbor is evaluated by
  running the robot with the newly chosen parameters. If this neighbor
  results in a longer distance walked than the previous best gait, it
  is saved as the new best gait. The process is then repeated, always
  starting with the best gait.

\item \emph{Gaussian random hill climbing}: This method works
  similarly to Uniform random hill climbing, except neighbors are
  generated by adding random Gaussian noise to the current best gait.
  This results in all parameters being changed at once, but the
  resulting vector is always fairly close to the previous best gait.
  We used independently selected noise in each dimension, scaled such
  that the standard deviation of the noise was 5\% of the range of
  that dimension.

\item \emph{N-dimensional policy gradient descent}: As opposed to the
  previous methods, this method explicitly estimates the gradient for
  the objective function. It does this by first evaluating \emph{t}
  randomly generated parameter vectors near the initial vector, each
  dimension of these vectors being perturbed by either $-\epsilon$,
  $0$, or $\epsilon$. Then, for each dimension, it groups vectors into
  three groups: $-\epsilon$, $0$, and $\epsilon$.  The gradient along
  this dimension is then estimated as the average score for the
  $\epsilon$ group minus the average score for the $-\epsilon$
  group. Finally, the method creates a new vector by changing all
  parameters by a fixed-size step in the direction of the gradient.

\item \emph{Nelder-Mead simplex method}\cite{nm}: The Nelder-Mead
  simplex method creates an initial simplex with 6 vertices. The
  initial parameter vector is stored as the first vertex and the other
  five vertices are created by adding to one dimension at a time one
  tenth of the allowable range for that parameter. It then tests the
  fitness of each vertex and based on these fitnesses, it reflects the
  worst point over the centroid in an attempt to improve it.  However,
  to prevent cycles and becoming stuck in local minima, several other
  rules are used.  In general, the worst vertex is reflected over the
  centroid. If the reflected point is better than the second worst
  point and worse than the best point, then the reflected point
  replaces the worst. If the reflected point is better than the best
  point, the simplex is expanded in the direction of the reflected
  point. The better of the reflected and the expanded point replaces
  the worst point. If the reflected point is worse than the second
  worst point, then the simplex is contracted away from the reflected
  point. If the contracted point is better than the reflected point,
  the contracted point replaces the worst point. If the contracted
  point is worse than the reflected point, the entire simplex is
  shrunk \cite{NMWebsite}.

\item \emph{Linear regression}: To initialize, this method chooses and
  evaluates five random parameter vectors. It then fits a linear model
  from parameter vector to fitness. In a loop, the method chooses and
  evaluates a new parameter vector generated by taking a fixed-size
  step in the direction of the gradient for each parameter, and fits a
  new linear model to all vectors evaluated so far, choosing the model
  to minimize the sum of squared errors.

\item \emph{SVM regression}: Similarly to linear regression, this
  model starts with several random vectors, but this time they are
  chosen in a small neighborhood about some initial random vector.
  These vectors (generally 8) are evaluated, and a support vector
  regression model is fit to the observed fitnesses.  To choose the
  next vector for evaluation, we randomly generate some number
  (typically 100) of vectors in the neighborhood of the best observed
  gait, and select for evaluation the vector with the best predicted
  performance.

  \edit{Has many parameters... in simulation... we found that it's
    better to bump points...}

\item \emph{Evolutionary Neural Network (HyperNEAT)\cite{clune}}

  \edit{do this}

@@

Various maching learning techniques have proven to be useful in
finding control policies for a wide variety of robots. Kohl and
Stone\cite{kohl} presented a policy gradient reinforcement learning
approach for generating a fast walk on legged robots. We experimented
with this method to create a walk for our robot (Policy Gradient
Descent). Chernova and Velosa\cite{chernova} took an evolutionary
approach to this problem which we did not implement.  Zykov, Bongard,
and Lipson\cite{zykov} describe the evolution of dynamic gaits on a
physical robot requiring no prior assumptions about the locomotion
pattern beyond the fact that it should be rhythmic.

@@

The quadruped robot has an on-board computer running Linux. The lower
level drivers are in C and the system is implemented in
Python. Feedback about distance travelled is provided via an infrared
LED mounted on the robot and a Wii remote fixed to the ceiling.

An overview of the code follows.

@@

\item \code{RunManager} class: Deals with all the details of running
  the robot, including running the robot multiple times, tracking
  distance walked via a \code{WiiTrackClient} member object, and
  writing results to the log file. Also includes the
  \code{explore\_dimensions} method to generate plots by varying each
  parameter independently.

\item \code{Strategy} class: The user has a choice between eight
  different learning strategies: Random search, uniform random hill
  climbing, Gaussian random hill climbing, N-dimensional policy
  gradient descent, Nelder-Mead simplex, linear regression/prediction,
  SVM regression/prediction, and HyperNEAT evolution.  Each strategy
  must derive from the base \code{Strategy} class and must implement
  two methods: \code{getNext}, for getting the next parameter vector
  to try, and \code{updateResults}, for communicating results of a run
  back to the strategy.

\item \code{MotionModel} class: We implemented several motion models,
  the main being the \code{SineModel5} class.  All motion models take
  as input a parameter vector and produce as output a motion model,
  which is simply a function mapping from time to nine motor
  positions.  The \code{SineModel5} model commands positions based on
  a sine wave, creating a periodic pattern. The parameters are:
  amplitude, wavelength, a multiplier for the inner vs outer motors,
  multiplier for left vs right motors, and multiplier for back vs
  front motors.  Other similar models were tested, including a seven
  parameter model which allowed sine waves shifted in time, but these
  were not used as extensively in our experiments as
  \code{SineModel5}.

\item \code{WiiTrackClient} and \code{WiiTrackServer} classes and
  hardware: A Wii remote tracks the location of the robot through an
  infrared LED mounted on top of the robot. A \code{WiiTrackServer} is
  run on the robot and continuously tracks its position using the
  CWiid library\cite{cwiid} to interface with the remote via
  bluetooth.  The \code{RunManager} then makes a
  \code{WiiTrackClient}, which connects via socket to the tracking
  server and requests position updates periodically.
  \code{RunManager} currently gets the robot's position at the
  beginning of each run and then again at the end and uses this to
  calculate the net change in position.

@@

We controlled our experiments from a computer that was connected via a
wireless ethernet to the robot. The robot collected data about
distance walked automatically on its own. If it walked outside of the
Wii remote's viewable area, it informed the user, so the only human
intervention required during an experiment was to move the robot back
inside the viewable area and to resume the run, which did not
interrupt the learning process or result in the loss of data.

We evaluated the efficacy of a set of parameters by sending these
parameters to the robot and instructing it to walk for a certain
length of time. The robot always began from the same position and
returned to the starting position at the end of the run in order to
measure true displacement without giving credit for ending in a leaned
position. More efficient parameters resulted in a faster gait, which
translated into a longer distance walked and a better score. After completing
an evaluation, the robot sent the resulting distance walked back to the
host computer and prepared itself for a new set of parameters to evaluate.

@@

The resulting gaits from our algorithms quickly outperformed the original
hand-coded walk designed for this robot. The fastest walk, for example, was
4 times faster.

@@

We have done complete runs of 3 different initial parameter vectors
for random search, uniform random hill climbing, Gaussian random hill
climbing, policy gradient descent, Nelder-Mead simplex, and linear
regression.  We also evaluated the SVM regression and HyperNEAT
methods, but these methods were more experimental, and thus we do not
yet have the same volume of data for these runs.  We developed several
gaits that were about 4 times faster than the original hand-coded
gait.

\edit{Convert to Table}

\begin{tabular}{|r|c|c|c||c|}
\hline
                                         & A       & B      & C      &  Average \\
\hline                                                               
\hline                                                               
Previous hand-coded gait                 & --      & --     & --     &  5.16 \\
\hline                                                                 
Random search                            & 6.04    & 17.26  & 4.90   &  9.40 \\
\hline                                                                 
Uniform Random Hill Climbing             & 11.37   & 9.44   & 2.69   &  7.83 \\
\hline                                                                 
Gaussian Random Hill Climbing            & 3.10    & 13.59  & 13.40  &  10.03 \\
\hline                                                                 
Policy Gradient Descent                  & 0.68    & 14.69  & 3.60   &  6.32 \\
\hline                                                                 
Nelder-Mead simplex                      & 8.51    & 13.62  & 14.83  &  12.32 \\
\hline                                                                 
Linear Regression                        & 27.58   & 12.51  & 1.95   &  14.01 \\
%\hline                                                                
%SVM Regression                           & ???     & ???    & ???   &   \\
%\hline                                                                
%Evolutionary Neural Network (HyperNEAT)  & ???     & ???    & ???   &   \\
\hline
\end{tabular}

\edit{Caption: The best gaits found for each starting vector and
  algorithm, in body lengths per minute.}

@@

\item For vector A, shown in \figref{vectorA}, linear regression
  worked significantly better than the other algorithms, resulting in
  a gait (27.58 body lengths/minute) that walked over twice as fast as
  the next best gait -- uniform random hill climbing at 11.37 body
  lengths/minute -- and 5.3 times better than the previous hand-coded
  gait.

\item Vector B, depicted in \figref{vectorB}, resulted in similarly
  performing gaits with all the algorithms. The random method got
  lucky and produced the best gait (17.26 body lengths/minute) and
  uniform random hill climbing produced the worst (9.44 body
  lengths/minute). The remaining algorithms each produced a gait
  around 13 body lengths/minute.

\item For vector C, shown in \figref{vectorC}, simplex and Gaussian
  random hill climbing each produced a gait that substantially
  outperformed the other algorithms. Simplex resulted in a gait of
  nearly 15 body lengths/minute and Gaussian random hill climbing
  produced a gait of just over 13 body lengths/minute, whereas the
  other algorithms returned gaits of less than 5 body lengths/minute.

@@

\figvarp{vectorA}{1}{Results for runs beginning with vector A.
  Linear regression worked significantly better than the other
  algorithms, resulting in a gait (27.58 body lengths/minute) that
  walked over twice as fast as the next best gait -- uniform random
  hill climbing at 11.37 body lengths/minute -- and 5.3 times better
  than the previous hand-coded gait.}{}

\figvarp{vectorB}{1}{Results for run starting with vector B.  All
  algorithms performed similarly.  The random method actually got
  lucky this time and produced the best gait (17.26 body
  lengths/minute) and uniform random hill climbing produced the worst
  (9.44 body lengths/minute). The remaining algorithms each produced a
  gait around 13 body lengths/minute.}{}

\figvarp{vectorC}{1}{Results for runs starting with vector C.
  Simplex and Gaussian random hill climbing each produced a gait that
  substantially outperformed the other algorithms, with around 15 body
  lengths/minute and 13 body lengths/minute, respectively, whereas the
  other algorithms returned gaits of less than 5 body
  lengths/minute.}{}

Based on the three trials, as shown in \figref{std_error}, linear
regression worked the best, followed by simplex. Gaussian random hill
climbing performed about as well as random, but it is unclear whether
Gaussian would continue to improve and become much better than random
search if allowed to run for more iterations. Uniform random hill
climbing and policy gradient descent performed similarly to each other
on average, but differed greatly on individual runs.

\figvarp{std_error}{1}{Average results for each method starting with
  parameter vectors A, B, and C.  Error bars are plotted showing the
  standard error.  Linear regression outperformed other methods,
  followed by simplex. Gaussian random hill climbing performed about
  as well as random, but it is unclear whether Gaussian would continue
  to improve and become much better than random search if allowed to
  run for more iterations. Uniform random hill climbing and policy
  gradient descent performed similarly to each other on average, but
  differed greatly on individual runs.}{}

In addition to running strategies to optimize parameterized gaits, we
also ......

@@

\edit{Jason: add exploration runs}

@@

Because only three trials were tested with each algorithm and no
algorithm consistently outperformed the others, there is a large
standard error for each algorithm, as show in \figref{std_error}. For
this reason, it is unclear if any algorithm is superior to another in
this application. More trials with each algorithm would be necessary
to reach a definitive ranking of the performance of the algorithms.

However, each algorithm discovered at least one gait of over 10 body
lengths/minute, including random. For this reason, we speculate that
the motion model is at least as critical as the learning algorithm
used.

@@

There are several directions in which we could continue our
work. First, the error margins for our runs were large, so reducing
them by running more trials would lead to more conclusive data. It
would also be insightful to run our algorithms on different motion
models.  We suspect that our choice of motion model influenced the
results greatly, as even random choices in the space produced gaits
that moved a significant fraction of the time (> 5\%).  It would be
interesting to see how learning methods would perform using a model
that included a much higher percentage of unproductive gaits. We also
intend to experiment more with evolutionary algorithms/HyperNEAT.
Some parameter vectors resulted in the robot turning, as opposed to it
walking long distances. This could be an interesting learning goal in
future projects. To these ends, we propose the following additions and
enhancements:

\begin{itemize}
\item More runs and/or longer runs
\item Different motion representations
\item Evolutionary algorithms/HyperNEAT\cite{clune}
\item Learning how to turn
\end{itemize}

@@

We have presented an array of approaches to optimizing a quadrupedal
gait for forward speed.  We have implemented and tested different
learning strategies, including uniform and Gaussian random hill
climbing, policy gradient reinforcement learning, Nelder-Mead simplex,
several new predictive methods based on linear and support vector
regression, and an evolved neural network (HyperNEAT).  We have also
compared these approaches to random search as a baseline. Many of the
methods resulted in walks significantly faster than previously
hand-tuned gaits.

Because only three trials were tested with each algorithm and no
algorithm consistently outperformed the others, there was a large
standard error for each method, as shown in \figref{std_error}. Thus
it was unclear if any algorithm was superior to another in this
application. More trials with each algorithm would be necessary to
reach a definitive ranking. Because each algorithm discovered at least one gait of
over 10 body lengths/minute, including random search, we also
conjectured that the motion representation for the robot is more
integral to forward speed than the learning algorithm.  How to learn
the motion representation, in addition to its parameters, remains an
open problem.

\section{Acknowledgments}
% List any people not on the team who helped you with your project: Your
% TA, other people you consulted with or had useful discussions (say in 
% a word or two what they did), people who proofread your report, and 
% any external code you used (libraries etc).
\begin{itemize}
\item Hod Lipson, Cornell Computational Synthesis Lab: adviser.
\item Jim T\o rreson, University of Oslo: adviser.
\item Juan Zagal, University of Chile: designed and printed robot and provided code for hand tuned gait.
\item Jeff Clune, CCSL Red Couch: collaborated on HyperNEAT implementation/testing.
\item Cooper Bills, Cornell University: assisted with Wii tracker development.
\item Anshumali Srivastava, Cornell University: Teaching Assistant.
\end{itemize}

\section{The Team}
% A table listing the names of the people who worked on the project, and who
% did what.

\item Diana Hidalgo: 
\begin{itemize}
\item Simplex (Nelder-Mead)
\item \code{WiiTrackClient} and \code{WiiTrackServer} classes and hardware
\item Plotting and evaluating results
\item Investigation into using Optitrack IR system for tracking robot
\end{itemize}

\item Sarah Nguyen
\begin{itemize}
\item Random search
\item Uniform random hill climbing
\item Gaussian random hill climbing
\item N-dimensional policy gradient descent
\item Linear regression and prediction
\item \code{RunManager} and \code{Strategy} classes
\item Video editing
\end{itemize}

\item Jason Yosinski
\begin{itemize}
\item SVM regression and prediction
\item Evolutionary Neural Network (HyperNEAT)
\item \code{WiiTrackClient} and \code{WiiTrackServer} classes and hardware
\item \code{Robot}, \code{RunManager}, and \code{Strategy} classes
\item \code{Parameterized Motion Model} abstract base class
\item \code{SineModel} classes
\item \code{explore\_dimensions}
\item Plotting and evaluating results
\end{itemize}

@@

\begin{itemize}
\item \code{optimize.py}: Main program, determines a strategy to try and runs the robot
  with that strategy.
\item \code{RunManager.py}: Deals with all the details of running the robot,
  including choosing an initial parameter, running the robot multiple times, 
  tracking distance walked, and writing to the log file. Also includes the
  \code{explore\_dimensions} method.
\item \code{Strategy.py}: Contains all the different possible strategies, which
  will be passed as objects in \code{optimize.py}. 
\item \code{Robot.py}: Implements the \code{Robot} class, described in
  \secref{implement}.
\item \code{SineModel.py}: Implements a sine based motion model,
  described in \secref{implement}.
\item \code{Motion.py}: Motion helper functions.
\item \code{WiiTrackServer.py}: Broadcasts the position of the infrared LED.
\item \code{WiiTrackClient.py}: Connects to the WiiTrackServer to get the current position information.
\end{itemize}
